Analysis of National Weather Storm Dataset
========================================================
Coursera - Reproducible Research: Peer Assessment Project #2

## Synopsis

## Data Processing

Supporting documentation can be found here:
- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

```{r}
dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dataBZ2 <- "StormData.csv.bz2"
#uncomment the following lines to retrieve the data
#download.file(dataURL, dataBZ2 )   #47MB
```

```{r}
SE <- read.csv(bzfile(dataBZ2)) #Storm Events
```
SE.casualty <- SE[SE$FATALITIES > 0 | SE$INJURIES > 0,]
SE.casualty$BGN_DATE <- as.Date(SE.casualty$BGN_DATE, "%m/%d/%Y")
SE.casualty$year <- as.POSIXlt(SE.casualty$BGN_DATE)$year +1900

library(lattice)
xyplot(FATALITIES ~ year, data=SE.casualty)
xyplot(FATALITIES ~ year | Event, data=SE.casualty)
library(plyr)
CbyEV <- ddply(SE.casualty,"EVTYPE", summarize,
             Fatalities=sum(FATALITIES,na.rm=TRUE ), 
             Injuries=sum(INJURIES, na.rm=TRUE))
CbyEV$Casualties <- CbyEV$Fatalities + CbyEV$Injuries

groupEvents <- function(df) {
  Event = as.numeric(nrow(df))                        #add event groups
  for (i in 1:nrow(df)) {
   if (grepl("TORN", df$EVTYPE[i], ignore.case=TRUE)) {
      Event[i] = "Tornado Related"
   } else if (grepl("COLD",  df$EVTYPE[i], ignore.case=TRUE)
           |  grepl("SNOW",  df$EVTYPE[i], ignore.case=TRUE) 
           |  grepl("ICE",   df$EVTYPE[i], ignore.case=TRUE) 
           |  grepl("FREEZ",   df$EVTYPE[i], ignore.case=TRUE) 
           |  grepl("WINTE",   df$EVTYPE[i], ignore.case=TRUE) 
           |  grepl("BLIZZ", df$EVTYPE[i], ignore.case=TRUE)) {
      Event[i] = "Cold Related"
   } else if (grepl("HEAT",  df$EVTYPE[i], ignore.case=TRUE)
           |  grepl("WARM",  df$EVTYPE[i], ignore.case=TRUE)) {
      Event[i] = "Heat Related"
   } else if (grepl("Flood",  df$EVTYPE[i], ignore.case=TRUE)) {
      Event[i] = "Flood Related"
   } else if (grepl("Thund",  df$EVTYPE[i], ignore.case=TRUE)) {
      Event[i] = "Thunderstorm Related"
   } else  {
      Event[i] = "Other"
   }
  }
}

for( iEV in sort(unique(CbyEV$Event))) {
   cat (iEV, ":\n")
   cat ( paste(sort(unique( as.character(CbyEV$EVTYPE[CbyEV$Event==iEV]))), sep=", "), "\n"  )
}

CbyEvent <- ddply(CbyEV,"Event", summarize, Fatalities=sum(Fatalities, na.rm=TRUE ))
FbyEvent
